---
title: "TBdar_fulldataset_morepopulations_statistics"
author: "Michaela Zwyer"
date: "2024-08-06"
output: html_document
---

# check, if any necessary packages are missing

```{r}
list.of.packages <- c("dplyr", "compositions", "finalfit", "stats", "lmtest")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```





```{r}
library(dplyr)
tbdar = read.delim("~/TBRU/Sinergia/Analysis/Human/Ancestry/Data/interim/metadata_Sinergia_final_all_pats_morePopulations.txt",
                  sep = "\t")



tbdar$Introduction <- ifelse(tbdar$Introduction %in% c("Intro 1", "Intro 5", "Intro 9", "Intro 10", "other"), tbdar$Introduction, NA)

tbdar$Intro10 <- ifelse(tbdar$Introduction == "Intro 10", 1, ifelse(tbdar$Introduction %in% c("Intro 1", "Intro 5", "Intro 9", "other"), 0, NA))
# create category for TB score
tbdar$TB_score_category <- ifelse(tbdar$TB_score < 6, "mild", ifelse(tbdar$TB_score > 5 & tbdar$TB_score < 8, "moderate", ifelse(tbdar$TB_score > 7, "severe", "somethingelse")))
tbdar$TB_score_2cat <- ifelse(tbdar$TB_score_category == "mild", "mild", ifelse(tbdar$TB_score_category %in% c("moderate", "severe"), "severe", NA))

tbdar$TB_category <- ifelse(tbdar$TB_category == "relapse", "relapse", ifelse(tbdar$TB_category == "new", "new", NA))
tbdar$Intro10 <- as.factor(tbdar$Intro10)
tbdar$DR_status <- ifelse(tbdar$DR_status %in% c("INH-Mono", "MDR", "Susceptible"), tbdar$DR_status, NA)
# age

tbdar$age_cat <- ifelse(tbdar$age < 30, 1, ifelse(tbdar$age >= 30 & tbdar$age < 40, 2, ifelse(tbdar$age >= 40 & tbdar$age < 50, 3, ifelse(tbdar$age >= 50, 4, NA))))

# add whether they are patients or control

tbdar$Type <- ifelse(tbdar$PATIENT_ID >= 83000 & !tbdar$Lineage %in% c("L1", "L2", "L4", "L3"), "control", "patient")

# make a column containing the common tribes
tbdar$common_tribes <- ifelse(unname(table(tbdar$patient_tribe)[tbdar$patient_tribe]) > 100, tbdar$patient_tribe, "other") # at least 100 patients with that tribe

tbdar$socio_cat <- ifelse(tbdar$socioeconomic_status < 100, 1, ifelse(tbdar$socioeconomic_status >= 100 & tbdar$socioeconomic_status < 200, 2, ifelse(tbdar$socioeconomic_status >= 200, 3, NA)))
# create a category for X-ray score:
tbdar$Xray_score_category_ralph <- ifelse(tbdar$Xray_score < 71, "mild", ifelse(tbdar$Xray_score >= 71, "severe", NA))
tbdar$Xray_01 <- ifelse(tbdar$Xray_score_category_ralph == "mild", 0, ifelse(tbdar$Xray_score_category_ralph == "severe", 1, NA))
unique(tbdar$Xray_score_category_ralph)
tbdar$Xray_score_category_ralph <- as.factor(tbdar$Xray_score_category_ralph)
# categorize the cough duration

tbdar$cough_cat <- ifelse(tbdar$symptoms_duration_cough_duration >= 4, 3, ifelse(tbdar$symptoms_duration_cough_duration < 4 & tbdar$symptoms_duration_cough_duration >= 3, 2, ifelse(tbdar$symptoms_duration_cough_duration < 3 & tbdar$symptoms_duration_cough_duration >= 0, 1, NA)))
# modify the empty HIV status to NA
tbdar$HIV_status <- ifelse(tbdar$HIV_status %in% c("infected", "positive"), "infected", ifelse(tbdar$HIV_status == "negative", "negative", NA))
unique(tbdar$HIV_status)
tbdar$TB_RF_smoking <- ifelse(tbdar$TB_RF_smoking == "no", "no", ifelse(tbdar$TB_RF_smoking == "yes", "yes", NA))
# add whether they are patients or control
# categories for age
tbdar$age_cat <- ifelse(tbdar$age < 30, 1, ifelse(tbdar$age >= 30 & tbdar$age < 40, 2, ifelse(tbdar$age >= 40 & tbdar$age < 50, 3, ifelse(tbdar$age >= 50, 4, NA))))

tbdar$DR_category <- ifelse(tbdar$DR_status %in% c("INH-Mono", "MDR"), "resistant", ifelse(tbdar$DR_status =="Susceptible", "susceptible", NA))
# turn all of them into a factor:
tbdar$TB_score_category <- as.factor(tbdar$TB_score_category)
tbdar$Xray_score_category <- as.factor(tbdar$Xray_score_category)
tbdar$patient_sex <- as.factor(tbdar$patient_sex)
tbdar$HIV_status <- as.factor(tbdar$HIV_status)
tbdar$TB_RF_smoking <- as.factor(tbdar$TB_RF_smoking)
tbdar$Xray_01 <- as.factor(tbdar$Xray_01)
tbdar$general_examination_malnutrition <- as.factor(tbdar$general_examination_malnutrition)
tbdar$Education <- as.factor(tbdar$Education)
tbdar$Type <- as.factor(tbdar$Type)
tbdar$DR_status <- as.factor(tbdar$DR_status)
tbdar$DR_category <- as.factor(tbdar$DR_category)
tbdar$age_cat <- as.factor(tbdar$age_cat)
tbdar$cough_cat <- as.factor(tbdar$cough_cat)
tbdar$TB_score_2cat <- as.factor(tbdar$TB_score_2cat)
tbdar$TB_category <- as.factor(tbdar$TB_category)
# relevel some factors:
tbdar$patient_sex <- relevel(tbdar$patient_sex, "male")
tbdar$HIV_status <- relevel(tbdar$HIV_status, "negative")
tbdar$Xray_01 <- relevel(tbdar$Xray_01, 1)
tbdar$TB_score_2cat <- relevel(tbdar$TB_score_2cat, "severe")
tbdar$DR_status <- relevel(tbdar$DR_status, "Susceptible")
tbdar$DR_category <- relevel(tbdar$DR_category, "susceptible")
tbdar$Education <- relevel(tbdar$Education, "primary") # most patients have primary education
tbdar$general_examination_malnutrition <- relevel(tbdar$general_examination_malnutrition, "no")
tbdar$other <- 1-(tbdar$Western_Bantu + tbdar$Southeastern_Bantu + tbdar$Eastern_Bantu)
tbdar$socio_cat <- as.factor(tbdar$socio_cat)
```

# look at the distribution of the continuous covariates you want to include

```{r}
# only include patients with either a bacterial or a human genome

tbdar_subset <- subset(tbdar, Introduction %in% c("other", "Intro 1", "Intro 5", "Intro 9", "Intro 10") | Nigerian > 0, select=c("age", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other",  "symptoms_duration_cough_duration", "socioeconomic_status"))

for (variable in c("age", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other",  "symptoms_duration_cough_duration", "socioeconomic_status")){
  hist(tbdar_subset[[variable]], main = variable, breaks = seq(min(tbdar_subset[[variable]], na.rm = T), max(tbdar_subset[[variable]], na.rm = T), length.out = 25)) 
}
```


# create the tables shown in the paper
# Table 1 

```{r}
# Table 1 Manuscript
library(finalfit)
library(readr)
# only include patients with either a bacterial or a human genome

tbdar_subset <- subset(tbdar, Introduction %in% c("other", "Intro 1", "Intro 5", "Intro 9", "Intro 10") | Nigerian > 0, select=c("Xray_01", "Introduction",  "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other"))

explanatory = c("Introduction", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu",  "other")
dependent = "Xray_01"
tbdar_subset %>%
    summary_factorlist(dependent = dependent, explanatory,  add_col_totals = T, add_row_totals = T, include_row_missing_col = T, p = F, na_include_dependent = TRUE, cont = "median") -> t1


t1

```


```{r}
# Table 1 Manuscript
library(finalfit)
library(readr)
# only include patients with either a bacterial or a human genome

tbdar_subset <- subset(tbdar, Introduction %in% c("other", "Intro 1", "Intro 5", "Intro 9", "Intro 10") | Nigerian > 0, select=c("TB_score_2cat", "Introduction",  "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other"))

explanatory = c("Introduction", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu",  "other")
dependent = "TB_score_2cat"
tbdar_subset %>%
    summary_factorlist(dependent = dependent, explanatory,  add_col_totals = T, add_row_totals = T, include_row_missing_col = T, p = F, na_include_dependent = TRUE) -> t1

t1

```
```{r}
# Table 1 Manuscript
library(finalfit)
library(readr)
# only include patients with either a bacterial or a human genome

tbdar_subset <- subset(tbdar, Introduction %in% c("other", "Intro 1", "Intro 5", "Intro 9", "Intro 10") | Nigerian > 0, select=c("Ct_value", "Introduction",  "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other"))

explanatory = c("Introduction", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu",  "other")
dependent = "Ct_value"
tbdar_subset %>%
    summary_factorlist(dependent = dependent, explanatory,  add_row_totals = T, include_row_missing_col = T,  p = F, na_include_dependent = TRUE) -> t1


t1

```

# Table 2

# add some more covariates
# calculate the median and IQR
```{r}
library(finalfit)
library(readr)
library(stats)
# calculate the 90% central range

central_range <- function(x)
  if (is.numeric(x)){
    cr <- quantile(x, probs = c(0.05, 0.95), na.rm = na.rm)
    return(sprintf("%.2f-%.2f", cr[1], cr[2]))
  } else {return(NA)}




# only include patients with either a bacterial or a human genome

tbdar_subset <- subset(tbdar, Introduction %in% c("other", "Intro 1", "Intro 5", "Intro 9", "Intro 10") | Nigerian > 0, select=c("Introduction", "patient_sex", "age", "HIV_status", "TB_RF_smoking", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu", "other",  "symptoms_duration_cough_duration", "socioeconomic_status", "Education", "general_examination_malnutrition", "TB_category", "DR_status"))

explanatory = c("patient_sex", "age", "HIV_status", "TB_RF_smoking", "symptoms_duration_cough_duration", "socioeconomic_status","Education", "general_examination_malnutrition", "TB_category", "DR_status", "Southeastern_Bantu", "Eastern_Bantu", "Western_Bantu",  "other")
dependent = "Introduction"
tbdar_subset %>%
    summary_factorlist(dependent = dependent, explanatory,  add_col_totals = T, add_row_totals = T, include_row_missing_col = T,  p = F, na_include_dependent = TRUE, cont = "median") -> t1



t1

```


# adjusting for the compositional variables and categorizing where necessary

```{r}
library(compositions)
# use the additive log ratio transformation to handle the ancestries since they are compositional variables (sum up to 1)

alr_data <- alr(tbdar[,c("Western_Bantu", "Southeastern_Bantu", "Eastern_Bantu","other")])
colnames(alr_data) <- c("Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr")
new_data <- cbind(tbdar, alr_data)

```


# build the model for X-ray score ending up with a model including the interaction
# from now on, exclude HIV-positive patients



# put the ancestries into bins and check, whether there is a linear relationship

```{r}
library(ggplot2)
library(mgcv)
for (anc in c("Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr")){
  hist(new_data[, anc], xlab = anc, main = "")
}
# make some categories

new_data$Western_alr_cat <- ifelse(new_data$Western_Bantu_alr >= -1, 4, ifelse(new_data$Western_Bantu_alr < -1 & new_data$Western_Bantu_alr > -2, 3, ifelse(new_data$Western_Bantu_alr <= -2 & new_data$Western_Bantu_alr > -6, 2,  ifelse(new_data$Western_Bantu_alr <= -6, 1, NA))))

new_data$Eastern_alr_cat <- ifelse(new_data$Eastern_Bantu_alr >= 0, 3, ifelse(new_data$Eastern_Bantu_alr < 0 & new_data$Eastern_Bantu_alr > -0.5, 2, ifelse(new_data$Eastern_Bantu_alr <= -0.5 , 1, NA)))
new_data$Southeastern_alr_cat <- ifelse(new_data$Southeastern_Bantu_alr >= 1, 3, ifelse(new_data$Southeastern_Bantu_alr < 1 & new_data$Southeastern_Bantu_alr > 0, 2, ifelse(new_data$Southeastern_Bantu_alr <= 0 , 1, NA)))


for (anc in c("Western", "Southeastern", "Eastern")){
  print(anc)
  print(table(new_data[,paste0(anc, "_alr_cat")]))
}


# how is the relationship between the outcome variables and the socioeconomic status?
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Xray_01", "socioeconomic_status")])
test <- glm(Xray_01 ~ socioeconomic_status, data = clean, na.action = na.exclude, family = binomial)
plot(test) # linearity looks fine to me according to https://bookdown.org/egarpor/PM-UC3M/glm-diagnostics.html Figure 5.14
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("TB_score_2cat", "socioeconomic_status")])
test <- glm(TB_score_2cat ~ socioeconomic_status, data = clean, na.action = na.exclude, family = binomial)
plot(test)
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Ct_value", "socioeconomic_status")])
test <- glm(log10(Ct_value) ~ socioeconomic_status, data = clean, na.action = na.exclude, family = gaussian)
plot(test)


```

# look at the model assumptions of the dependent variable X-ray score

```{r}
library(car)
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "patient_sex", "age", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_score",  "symptoms_duration_cough_duration", "socioeconomic_status", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(Xray_score ~ Eastern_Bantu_alr + Southeastern_Bantu_alr + Western_Bantu_alr+ Intro10 + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = gaussian, na.action = na.exclude)
plot(fitted(test), resid(test))
abline(h=0, col = "red")
plot(test, 1) # look at the residuals vs fitted
durbinWatsonTest(test) # check, whether the predictors are independent and observed with negligible error
plot(test,3) # check, whether the residual erros have constant variance
```
# when categorizing the continuous covariates

```{r}
library(car)
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_score", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(Xray_score ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = gaussian, na.action = na.exclude)
plot(fitted(test), resid(test))
abline(h=0, col = "red")
plot(test, 1) # look at the residuals vs fitted
durbinWatsonTest(test) # check, whether the predictors are independent and observed with negligible error
plot(test,3) # check, whether the residual erros have constant variance
```
# when categorizing the Xray-score, too

```{r}
library(car)
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_01", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = binomial, na.action = na.exclude)
plot(fitted(test), resid(test))
abline(h=0, col = "red")
plot(test, 1) # look at the residuals vs fitted
durbinWatsonTest(test) # check, whether the predictors are independent and observed with negligible error
plot(test,3) # check, whether the residual erros have constant variance
```

# make some test, to check, whether the assumptions of the model are fulfilled

```{r}
# test the linearity on the log-odds scale
# do a Box-Tidwell test to see, whether the log transformed continuous variables have a significant interaction with themselves

library(car)

# X-ray score
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_01", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socioeconomic_status", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socioeconomic_status + socioeconomic_status:log(socioeconomic_status) + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = binomial, na.action = na.exclude)
logodds <- test$linear.predictors
summary(test)

# TB-score
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "TB_score_2cat", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socioeconomic_status", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socioeconomic_status + socioeconomic_status:log(socioeconomic_status) + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = binomial, na.action = na.exclude)
logodds <- test$linear.predictors
summary(test)

# Ct-value - look at the residual plots
clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Ct_value", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socioeconomic_status", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socioeconomic_status + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = "gaussian", na.action = na.exclude)
plot(test)

```

--> for TB-score, there is linearity with the socioeconomic status. Thus let us categorize the variable
--> added to the beginning already



# any multicollinearity between the explanatory variables?
--> seems fine, all VIF values are low...

```{r}
library(car)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_01", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + Education + TB_category + DR_category, data = clean, family = binomial, na.action = na.exclude)
vif(test)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "TB_score_2cat", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + Education + TB_category+ DR_category, data = clean, family = binomial, na.action = na.exclude)
vif(test)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "TB_score_2cat", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
test <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + Education + TB_category+ DR_category, data = clean, family = binomial, na.action = na.exclude)
vif(test)


```


# create a model and look at the relationship

```{r}
library(lmtest)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_01", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "general_examination_malnutrition", "TB_category", "Education", "DR_category", "DR_status")])

dim(clean)
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Southeastern_alr_cat) + as.factor(Eastern_alr_cat), data = clean, family = binomial, na.action = na.exclude)
print(summary(model1))
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(Xray_01 ~ 1, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
```

--> looking at the summary I would say that Western_Bantu and Southeastern_Bantu are not linear, thus use them as categories

```{r}
# add the introduction
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```
```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one
# sex

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# smoking

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# cough duration

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat) + general_examination_malnutrition, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```
```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat) + general_examination_malnutrition + TB_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```
```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat)+ general_examination_malnutrition + TB_category + Education, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat)+ general_examination_malnutrition + TB_category + Education + DR_status, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

--> there are too little MDR cases, categorize the DR-status into susceptible and resistant
--> adjusted in the beginning...


```{r}
# add the covariates


model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + as.factor(socio_cat)+ general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

# Compare to a model without the ancestries
# having more  covariables
# Table 3_revisions_morecovariates

```{r}
library(lmtest)
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking +  socio_cat + TB_category + general_examination_malnutrition + Education+ DR_category, data = clean, family = binomial, na.action = na.exclude)
model2 <- glm(Xray_01 ~ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking +  socio_cat+ TB_category + general_examination_malnutrition + Education+ DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# add the interaction

# with more covariates for table 3 revisions_morecovariates
```{r}
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking +  socio_cat + TB_category + general_examination_malnutrition + Education+ DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(Xray_01 ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking  + socio_cat + TB_category + general_examination_malnutrition + Education+ DR_category, data = clean, family = binomial, na.action = na.exclude)

lrtest(model2, model1) 
```

# does adding of cough duration have an effect?
--> not really, values change slightly...

```{r}
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + socio_cat + TB_category + general_examination_malnutrition + Education+ DR_category+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(Xray_01 ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking  + socio_cat + TB_category + general_examination_malnutrition + Education+ DR_category+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)

lrtest(model2, model1) 
```




# now for TB-score

# create a model and look at the relationship

```{r}
library(lmtest)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "TB_score_2cat", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat", "socio_cat", "TB_category", "general_examination_malnutrition", "Education", "DR_category")])
dim(clean)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Southeastern_alr_cat) + as.factor(Eastern_alr_cat), data = clean, family = binomial, na.action = na.exclude)
print(summary(model1))
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(TB_score_2cat ~ 1, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
```

```{r}
# add the introduction
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# age
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one
# sex

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# smoking

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# cough duration

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration + socioeconomic_status

# do it one by one
# socioeconomic_status

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


```{r}
# add the covariates

# do it one by one
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```
```{r}
# add the covariates

# do it one by one
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates

# do it one by one
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category + Education, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates

# do it one by one
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

# pvalue for the ancestries

# same with more covariates

```{r}
library(lmtest)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+  socio_cat+ general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)

model2 <- glm(TB_score_2cat ~ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+  socio_cat+ general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# add the interaction

# with more covariates

```{r}
library(lmtest)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+  socio_cat + general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
# compare to a model without the interaction
model2 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat)+ as.factor(Southeastern_alr_cat) + Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ socio_cat+ general_examination_malnutrition + TB_category + Education + DR_category, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# does the addition of the cough duration variable change the results?
--> no it does not


```{r}
library(lmtest)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+  socio_cat + general_examination_malnutrition + TB_category + Education + as.factor(cough_cat) +DR_category, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
# compare to a model without the interaction
model2 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat)+ as.factor(Southeastern_alr_cat) + Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ socio_cat+ general_examination_malnutrition + TB_category + Education + as.factor(cough_cat) +DR_category, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```


# now for the Ct-value

# Check, whether the assumptions for a glm are met

```{r}
# create a clean dataset with no missing values for any of the ancestries, bacterial genotype, and the covariates included and it only comprises HIV negative patients

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Ct_value", "cough_cat", "socio_cat", "general_examination_malnutrition", "TB_category", "Education", "DR_category")])

dim(clean)
# Linearity
# for the ancestries as category
model1 <- glm(Ct_value ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


par(mfrow = c(1,3))
for (anc in c("Western_alr_cat", "Eastern_alr_cat", "Southeastern_alr_cat")){
boxplot(residuals(model1) ~ clean[[anc]], main = paste("Residuals vs ", anc))
abline(h = 0, col = "red")}

# Homoscedasticity

plot(model1, which = 1)

# Normality of Residuals
# QQ-Plot

qqnorm(residuals(model1))
qqline(residuals(model1), col = "red")

# Histogram of residuals

hist(residuals(model1), main = "Histogram of residuals", xlab = "Residuals")

```

--> the qq-plot does not look good I would say.

# try a log transformation of the ct_value

```{r}
# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


par(mfrow = c(1,3))
for (anc in c("Western_alr_cat", "Eastern_alr_cat", "Southeastern_alr_cat")){
boxplot(residuals(model1) ~ clean[[anc]], main = paste("Residuals vs ", anc))
abline(h = 0, col = "red")}

# Homoscedasticity

plot(model1, which = 1)

# Normality of Residuals
# QQ-Plot

qqnorm(residuals(model1))
qqline(residuals(model1), col = "red")

# Histogram of residuals

hist(residuals(model1), main = "Histogram of residuals", xlab = "Residuals")

```
--> The qq-plot looks much better

# continue with log10 transformation 

```{r}
library(lmtest)

# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


summary(model1)

```
# add the introduction

```{r}
library(lmtest)

# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10, data = clean, na.action = na.exclude, family = "gaussian")

model2 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

lrtest(model1, model2)

```

# add the covariates one by one

```{r}
library(lmtest)

# sex
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)


```

```{r}
library(lmtest)

# age
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# smoking
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# cough duration
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# socioeconomic status
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat) + socio_cat, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# socioeconomic status
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# socioeconomic status
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# socioeconomic status
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category + Education, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# socioeconomic status
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + as.factor(cough_cat) + socio_cat + general_examination_malnutrition + TB_category + Education + DR_category, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

# test, whether a model including the ancestries is better than a model without human ancestry
# all ancestries

# add more covariates

```{r}
library(lmtest)

# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + socio_cat + TB_category + general_examination_malnutrition + Education + DR_category, data = clean, na.action = na.exclude, family = "gaussian")
model2 <- glm(log10(Ct_value) ~ Intro10 + patient_sex + as.factor(age_cat) + TB_RF_smoking + socio_cat+ TB_category + general_examination_malnutrition + Education +DR_category, data = clean, na.action = na.exclude, family = "gaussian")
lrtest(model1, model2)
summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# add the interaction 

# with more covariates

```{r}
library(lmtest)

# cough duration
# for the ancestries as category
model2 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking +  socio_cat + TB_category + general_examination_malnutrition + Education + DR_category, data = clean, na.action = na.exclude, family = "gaussian")
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)*Intro10+ as.factor(Eastern_alr_cat)*Intro10 + as.factor(Southeastern_alr_cat)*Intro10 + patient_sex + age_cat + TB_RF_smoking + socio_cat+ TB_category + general_examination_malnutrition + Education + DR_category, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
lrtest(model1, model2)

```

# does the addition of the cough duration variable change the results?
--> no it does not

```{r}
library(lmtest)

# cough duration
# for the ancestries as category
model2 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking + socio_cat + TB_category + general_examination_malnutrition + Education + DR_category + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)*Intro10+ as.factor(Eastern_alr_cat)*Intro10 + as.factor(Southeastern_alr_cat)*Intro10 + patient_sex + age_cat + TB_RF_smoking + socio_cat+ TB_category + general_examination_malnutrition + Education + DR_category + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
lrtest(model1, model2)

```
