---
title: "TBdar_fulldataset_morepopulations_statistics"
author: "Michaela Zwyer"
date: "2024-08-06"
output: html_document
---

# check, if any necessary packages are missing

```{r}
list.of.packages <- c("dplyr", "compositions", "finalfit", "stats", "lmtest")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```


```{r}
library(dplyr)
tbdar = read.delim("~/Sinergia/Analysis/Human/Ancestry/Data/interim/metadata_Sinergia_final_all_pats_morePopulations.txt",
                  sep = "\t")
head(tbdar)
unique(tbdar$age_category)
# create category for TB score
tbdar$TB_score_category <- ifelse(tbdar$TB_score < 6, "mild", ifelse(tbdar$TB_score > 5 & tbdar$TB_score < 8, "moderate", ifelse(tbdar$TB_score > 7, "severe", "somethingelse")))
unique(tbdar$TB_score_category)

# create a category for X-ray score:
tbdar$Xray_score_category_ralph <- ifelse(tbdar$Xray_score < 71, "mild", ifelse(tbdar$Xray_score >= 71, "severe", NA))
tbdar$Xray_01 <- ifelse(tbdar$Xray_score_category_ralph == "mild", 0, ifelse(tbdar$Xray_score_category_ralph == "severe", 1, NA))
unique(tbdar$Xray_score_category_ralph)
tbdar$Xray_score_category_ralph <- as.factor(tbdar$Xray_score_category_ralph)

# modify the empty HIV status to NA
tbdar$HIV_status <- ifelse(tbdar$HIV_status %in% c("infected", "positive"), "infected", ifelse(tbdar$HIV_status == "negative", "negative", NA))
unique(tbdar$HIV_status)
tbdar$TB_RF_smoking <- ifelse(tbdar$TB_RF_smoking == "no", "no", ifelse(tbdar$TB_RF_smoking == "yes", "yes", NA))



tbdar$Intro10 <- ifelse(tbdar$Introduction == "Intro 10", 1, ifelse(tbdar$Introduction %in% c("Intro 1", "Intro 5", "Intro 9", "other"), 0, NA))

tbdar$TB_score_2cat <- ifelse(tbdar$TB_score_category == "mild", "mild", ifelse(tbdar$TB_score_category %in% c("moderate", "severe"), "severe", NA))
tbdar$TB_score_2cat <- as.factor(tbdar$TB_score_2cat)

tbdar$Intro10 <- as.factor(tbdar$Intro10)
# turn all of them into a factor:
tbdar$TB_score_category <- as.factor(tbdar$TB_score_category)
tbdar$Xray_score_category <- as.factor(tbdar$Xray_score_category)
tbdar$patient_sex <- as.factor(tbdar$patient_sex)
tbdar$HIV_status <- as.factor(tbdar$HIV_status)
tbdar$TB_RF_smoking <- as.factor(tbdar$TB_RF_smoking)
tbdar$Xray_01 <- as.factor(tbdar$Xray_01)

# relevel some factors:
tbdar$patient_sex <- relevel(tbdar$patient_sex, "male")
tbdar$HIV_status <- relevel(tbdar$HIV_status, "negative")
tbdar$Xray_01 <- relevel(tbdar$Xray_01, 1)
tbdar$TB_score_2cat <- relevel(tbdar$TB_score_2cat, "severe")

```

# adjusting for the compositional variables and categorizing where necessary

```{r}
library(compositions)
# use the additive log ratio transformation to handle the ancestries since they are compositional variables (sum up to 1)

tbdar$other <- 1-(tbdar$Western_Bantu + tbdar$Southeastern_Bantu + tbdar$Eastern_Bantu)

alr_data <- alr(tbdar[,c("Western_Bantu", "Southeastern_Bantu", "Eastern_Bantu","other")])
colnames(alr_data) <- c("Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr")
new_data <- cbind(tbdar, alr_data)

```


# build the model for X-ray score ending up with a model including the interaction
# from now on, exclude HIV-positive patients

# put the ancestries into bins and check, whether there is a linear relationship

```{r}
for (anc in c("Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr")){
  hist(new_data[, anc], xlab = anc, main = "")
}
# make some categories

new_data$Western_alr_cat <- ifelse(new_data$Western_Bantu_alr >= -1, 4, ifelse(new_data$Western_Bantu_alr < -1 & new_data$Western_Bantu_alr > -2, 3, ifelse(new_data$Western_Bantu_alr <= -2 & new_data$Western_Bantu_alr > -6, 2,  ifelse(new_data$Western_Bantu_alr <= -6, 1, NA))))

new_data$Eastern_alr_cat <- ifelse(new_data$Eastern_Bantu_alr >= 0, 3, ifelse(new_data$Eastern_Bantu_alr < 0 & new_data$Eastern_Bantu_alr > -0.5, 2, ifelse(new_data$Eastern_Bantu_alr <= -0.5 , 1, NA)))
new_data$Southeastern_alr_cat <- ifelse(new_data$Southeastern_Bantu_alr >= 1, 3, ifelse(new_data$Southeastern_Bantu_alr < 1 & new_data$Southeastern_Bantu_alr > 0, 2, ifelse(new_data$Southeastern_Bantu_alr <= 0 , 1, NA)))
# age
hist(new_data$age)
new_data$age_cat <- ifelse(new_data$age < 30, 1, ifelse(new_data$age >= 30 & new_data$age < 40, 2, ifelse(new_data$age >= 40 & new_data$age < 50, 3, ifelse(new_data$age >= 50, 4, NA))))

for (anc in c("Western", "Southeastern", "Eastern")){
  print(anc)
  print(table(new_data[,paste0(anc, "_alr_cat")]))
}

# categorize the cough duration

hist(new_data$symptoms_duration_cough_duration)
new_data$cough_cat <- ifelse(new_data$symptoms_duration_cough_duration >= 4, 3, ifelse(new_data$symptoms_duration_cough_duration < 4 & new_data$symptoms_duration_cough_duration >= 3, 2, ifelse(new_data$symptoms_duration_cough_duration < 3 & new_data$symptoms_duration_cough_duration >= 0, 1, NA)))
table(new_data$cough_cat, new_data$Xray_01, exclude = NULL)
```

# create a model and look at the relationship

```{r}
library(lmtest)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Xray_01", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr")])

dim(clean)
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Southeastern_alr_cat) + as.factor(Eastern_alr_cat), data = clean, family = binomial, na.action = na.exclude)
print(summary(model1))
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(Xray_01 ~ 1, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
```

--> looking at the summary I would say that Western_Bantu and Southeastern_Bantu are not linear, thus use them as categories

```{r}
# add the introduction
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```
```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one
# sex

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# smoking

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# cough duration

model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


# Compare to a model without the ancestries


```{r}
library(lmtest)
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
model2 <- glm(Xray_01 ~ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```


# add the interaction


```{r}
model1 <- glm(Xray_01 ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(Xray_01 ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)

lrtest(model2, model1) 
```


# now for TB-score

# create a model and look at the relationship

```{r}
library(lmtest)

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "TB_score_2cat", "Western_Bantu_alr", "Southeastern_Bantu_alr", "Eastern_Bantu_alr", "cough_cat")])
dim(clean)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Southeastern_alr_cat) + as.factor(Eastern_alr_cat), data = clean, family = binomial, na.action = na.exclude)
print(summary(model1))
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
model2 <- glm(TB_score_2cat ~ 1, data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
```

```{r}
# add the introduction
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one
# age
model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + HIV_status + symptoms_duration_cough_duration

# do it one by one
# sex

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# smoking

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking, data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```

```{r}
hist(new_data[new_data$HIV_status == "negative" & new_data$Intro10 %in% c(0,1), ]$TB_score)
```
--> the distribution of TB_score looks quite normal, could we use it as continous variable instead of making it categorical?

```{r}
# add the covariates
# + age + patient_sex + TB_RF_smoking + symptoms_duration_cough_duration

# do it one by one
# cough duration

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat)+ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking + as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
```


# add the interaction


```{r}
library(lmtest)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat)* Intro10 + as.factor(Eastern_alr_cat) * Intro10 + as.factor(Southeastern_alr_cat)* Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
# compare to a model without the interaction
model2 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat)+ as.factor(Southeastern_alr_cat) + Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# pvalue for the ancestries


```{r}
library(lmtest)

model1 <- glm(TB_score_2cat ~ as.factor(Western_alr_cat) + as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)

model2 <- glm(TB_score_2cat ~ Intro10 + as.factor(age_cat) + patient_sex + TB_RF_smoking+ as.factor(cough_cat), data = clean, family = binomial, na.action = na.exclude)
summary(model1)
lrtest(model1, model2)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```

# now for the Ct-value

# Check, whether the assumptions for a glm are met

```{r}
# create a clean dataset with no missing values for any of the ancestries, bacterial genotype, and the covariates included and it only comprises HIV negative patients

clean <- na.omit(new_data[new_data$HIV_status == "negative", c("Western_alr_cat", "Southeastern_alr_cat", "Eastern_alr_cat", "patient_sex", "age_cat", "symptoms_duration_cough_duration", "Intro10", "TB_RF_smoking", "Ct_value", "cough_cat")])

dim(clean)
# Linearity
# for the ancestries as category
model1 <- glm(Ct_value ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


par(mfrow = c(1,3))
for (anc in c("Western_alr_cat", "Eastern_alr_cat", "Southeastern_alr_cat")){
boxplot(residuals(model1) ~ clean[[anc]], main = paste("Residuals vs ", anc))
abline(h = 0, col = "red")}

# Homoscedasticity

plot(model1, which = 1)

# Normality of Residuals
# QQ-Plot

qqnorm(residuals(model1))
qqline(residuals(model1), col = "red")

# Histogram of residuals

hist(residuals(model1), main = "Histogram of residuals", xlab = "Residuals")

```

--> the qq-plot does not look good I would say.

# try a log transformation of the ct_value

```{r}
# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


par(mfrow = c(1,3))
for (anc in c("Western_alr_cat", "Eastern_alr_cat", "Southeastern_alr_cat")){
boxplot(residuals(model1) ~ clean[[anc]], main = paste("Residuals vs ", anc))
abline(h = 0, col = "red")}

# Homoscedasticity

plot(model1, which = 1)

# Normality of Residuals
# QQ-Plot

qqnorm(residuals(model1))
qqline(residuals(model1), col = "red")

# Histogram of residuals

hist(residuals(model1), main = "Histogram of residuals", xlab = "Residuals")

```
--> The qq-plot looks much better

# continue with log10 transformation 

```{r}
library(lmtest)

# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")


summary(model1)

```
# add the introduction

```{r}
library(lmtest)

# Linearity
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10, data = clean, na.action = na.exclude, family = "gaussian")

model2 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

lrtest(model1, model2)

```

# add the covariates one by one

```{r}
library(lmtest)

# sex
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)


```

```{r}
library(lmtest)

# age
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# smoking
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking, data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

```{r}
library(lmtest)

# cough duration
# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)

```

# test, whether a model including the ancestries is better than a model without human ancestry
# all ancestries
```{r}
library(lmtest)

# for the ancestries as category
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")
model2 <- glm(log10(Ct_value) ~ Intro10 + patient_sex + age_cat + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")
lrtest(model1, model2)
summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
```


# add the interaction 

```{r}
library(lmtest)

# cough duration
# for the ancestries as category
model2 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)+ as.factor(Eastern_alr_cat) + as.factor(Southeastern_alr_cat) + Intro10 + patient_sex + age_cat + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")
model1 <- glm(log10(Ct_value) ~ as.factor(Western_alr_cat)*Intro10+ as.factor(Eastern_alr_cat)*Intro10 + as.factor(Southeastern_alr_cat)*Intro10 + patient_sex + age_cat + TB_RF_smoking + as.factor(cough_cat), data = clean, na.action = na.exclude, family = "gaussian")

summary(model1)
ORs <- exp(cbind(OR = coef(model1), confint(model1)))
round(ORs, digits = 2)
lrtest(model1, model2)

```



